---
title: "Basic NLGMM example"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
  pdf_document:
    toc: true
fig_caption: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE}
set.seed(17070523)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Nonlinear GLMM example

This example simulates nonlinear psychometric data and then fits models including a Bayesian nonlinear model using `brms`.

## Simulate data

We construct a dataset with stimulus intensity, stimulus type, and a binomial response.

```{r simulate-nlgmm}
n_animals <- 10
n_levels <- 7
n_types <- 2
replicates <- 20

width_alpha <- 0.8
width_coef <- 2 * log(2 / (1 - width_alpha) - 1)
inflex <- 15
width <- -10
lapse <- 0.12
base <- 0.5

pop_inflex <- 2.00
pop_sd <- 1.00
pop_width <- 3.00
width_sd <- 1.03

type_inflex <- 2.50
type_width <- 3.5
type_animal_sd <- 0.55
lapse_animal_sd <- 0.50
extra_error_sd <- 0.10

xx <- seq(from = 0, to = n_levels - 1, length.out = 1e3)
plot(x = xx,
     y = base + (1 - lapse - base) * plogis(q = width_coef * (xx - pop_inflex) / pop_width),
     xlim = c(0, n_levels - 1),
     ylim = c(0, 1),
     type = 'l',
     col = 'blue',
     xlab = 'Stimulus',
     ylab = 'Proportion correct')
abline(h = c(0, 1, base, 1 - lapse), lty = c(1, 1, 3, 3))

animal <- rep(x = 1:n_animals, times = n_levels)
animal <- sort(animal)
animal <- rep(x = animal, times = n_types)
stimulus <- rep(x = 1:n_levels - 1, times = n_animals)
stimulus <- rep(x = stimulus, times = n_types)
stype <- rep(x = 1:n_types, times = n_animals * n_levels)
stype <- sort(stype)

animal <- rep(x = animal, times = replicates)
stimulus <- rep(x = stimulus, times = replicates)
stype <- rep(x = stype, times = replicates)

base_bias_animal <- rnorm(n = n_animals, mean = pop_inflex, sd = pop_sd)
width_animal <- rnorm(n = n_animals, mean = pop_width, sd = width_sd)
type_animal <- rnorm(n = n_animals, mean = 1.0, sd = type_animal_sd)
lapse_animal <- plogis(rnorm(n = n_animals, mean = qlogis(lapse), sd = lapse_animal_sd))

response_y <- rep(x = NA, length = n_animals * n_levels * n_types * replicates)
for (ii in seq_along(response_y)) {
  response_y[ii] <- base + (1 - lapse_animal[animal[ii]] - base) *
    plogis(q = width_coef * (stimulus[ii] - (base_bias_animal[animal[ii]] + (stype[ii] - 1) * type_inflex)) /
             (width_animal[animal[ii]] + (stype[ii] - 1) * type_width * type_animal[animal[ii]]) +
             rnorm(n = 1, mean = 0, sd = extra_error_sd))
}

correct_incorrect <- rbinom(n = length(response_y), size = 1, prob = response_y)

sim_data <- data.frame(
  correct_incorrect = correct_incorrect,
  animal = LETTERS[animal],
  stimulus = stimulus,
  type = factor(stype, labels = c('alpha', 'beta'))
)

summary(sim_data)
```

## Save and load data

```{r save-read-nlgmm}
root_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), 'Simulated Data')
file_name <- 'simulated_nonlinear_data.csv'
write.csv(file = file.path(root_dir, file_name), x = sim_data, row.names = FALSE)

dta <- read.csv(file = file.path(root_dir, file_name), header = TRUE)
dta <- within(dta, {
  animal <- factor(animal)
  type <- factor(type)
})

summary(dta)
```

## Plot the data

We inspect the aggregated proportion correct by stimulus and type.

```{r plot-nlgmm, fig.width=7, fig.height=5}
clz <- sapply(length(unique(dta$animal)), colorRampPalette(c('red', 'blue', 'cyan4', 'magenta4', 'orange')))
clz <- sample(clz)

agg <- aggregate(correct_incorrect ~ stimulus * type * animal, data = dta, FUN = mean)
head(agg)

plot(NULL,
     xlab = 'Stimulus intensity',
     ylab = 'Proportion correct',
     xlim = range(stimulus) * c(0.7, 1.3),
     ylim = c(0, 1))
abline(h = c(0, 1, base, 1 - lapse), lty = c(1, 1, 3, 3))
for (aa in unique(agg$animal)) {
  with(subset(agg, animal == aa), {
    points(
      x = stimulus,
      y = correct_incorrect,
      col = clz[which(LETTERS %in% aa)],
      pch = c(20, 21)[1 + type %in% levels(type)[2]]
    )
  })
}
legend('topright', legend = unique(agg$animal), col = clz, pch = 20)
legend('bottomright', legend = unique(agg$type), pch = c(20, 21))
```

## Install and set up Bayesian modelling packages

This section describes package installation steps for `cmdstanr` and `brms`.

```{r setup-bayesian, eval=FALSE}
if (!any(installed.packages() %in% 'remotes')) install.packages('remotes')
if (!any(installed.packages() %in% 'cmdstanr')) remotes::install_github('stan-dev/cmdstanr')
require(cmdstanr)
install_cmdstan(cores = parallel::detectCores() - 1, overwrite = FALSE)
if (!any(installed.packages() %in% 'brms')) install.packages('brms')
require(cmdstanr)
require(brms)
```

## Define the nonlinear model

We specify a nonlinear formula with transformed parameters for baseline, lapse, inflection, and width.

```{r define-model, message=FALSE}
library(brms)
inv_logit <- inv_logit_scaled

formula_nl <- bf(
  correct_incorrect ~ inv_logit(LogitBase) + (1 - inv_logit(LogitLapse) - inv_logit(LogitBase)) *
    inv_logit(4.39 * (stimulus - Inflex) / exp(LogWidth)),
  LogitBase ~ 1,
  LogitLapse ~ type + (1 + type | animal),
  Inflex ~ type + (1 + type | animal),
  LogWidth ~ type + (1 + type | animal),
  family = bernoulli('identity'),
  nl = TRUE
)

formulaNull_nl <- bf(
  correct_incorrect ~ inv_logit(LogitBase) + (1 - inv_logit(LogitLapse) - inv_logit(LogitBase)) * inv_logit(LogitMean),
  LogitBase ~ 1,
  LogitLapse ~ 1 + (1 | animal),
  LogitMean ~ 1 + (1 | animal),
  family = bernoulli('identity'),
  nl = TRUE
)

prior_nl <- get_prior(formula = formula_nl, data = dta)
priorNull_nl <- get_prior(formula = formulaNull_nl, data = dta)
```

## Set custom priors

We replace default priors with more informative priors for the nonlinear parameters.

```{r priors}
prior_nl <- within(prior_nl, {
  prior[class %in% 'b' & nlpar %in% 'LogitBase' & coef %in% 'Intercept'] <- 'normal(0,1)'
  prior[class %in% 'b' & nlpar %in% 'LogitLapse' & coef %in% 'Intercept'] <- 'normal(-3,3)'
  prior[class %in% 'b' & nlpar %in% 'LogitLapse' & coef %in% 'typebeta'] <- 'normal(0,3)'
  prior[class %in% 'b' & nlpar %in% 'Inflex' & coef %in% 'Intercept'] <- 'normal(3,3)'
  prior[class %in% 'b' & nlpar %in% 'Inflex' & coef %in% 'typebeta'] <- 'normal(0,3)'
  prior[class %in% 'b' & nlpar %in% 'LogWidth' & coef %in% 'Intercept'] <- 'normal(1,3)'
  prior[class %in% 'b' & nlpar %in% 'LogWidth' & coef %in% 'typebeta'] <- 'normal(0,3)'
  prior
})

priorNull_nl <- within(priorNull_nl, {
  prior[class %in% 'b' & nlpar %in% 'LogitBase' & coef %in% 'Intercept'] <- prior_nl$prior[prior_nl$class %in% 'b' & prior_nl$nlpar %in% 'LogitBase' & prior_nl$coef %in% 'Intercept']
  prior[class %in% 'b' & nlpar %in% 'LogitLapse' & coef %in% 'Intercept'] <- prior_nl$prior[prior_nl$class %in% 'b' & prior_nl$nlpar %in% 'LogitLapse' & prior_nl$coef %in% 'Intercept']
  prior[class %in% 'b' & nlpar %in% 'LogitLapse' & coef %in% 'typebeta'] <- prior_nl$prior[prior_nl$class %in% 'b' & prior_nl$nlpar %in% 'LogitLapse' & prior_nl$coef %in% 'typebeta']
  prior[class %in% 'b' & nlpar %in% 'LogitMean' & coef %in% 'Intercept'] <- 'normal(0,3)'
  prior
})

print(prior_nl)
print(priorNull_nl)
```

## Fit the model

The Bayesian model is fit in stages: a prior-only dummy run, a short run, and a full run.

```{r fit-model, eval=FALSE}
# Dummy prior-only fit
dummy_fit <- brm(formula = formula_nl, data = dta, prior = prior_nl, sample_prior = 'only', iter = 300, chains = 4, cores = 4, refresh = 0, backend = 'cmdstanr')
plot(dummy_fit, nvariables = 10, variable = '^b_', regex = TRUE)
plot(dummy_fit, nvariables = 10, variable = '^sd', regex = TRUE)

# Short fit for convergence check
short_fit <- brm(formula = formula_nl, data = dta, prior = prior_nl, iter = 300, chains = 4, cores = 4, refresh = 0, backend = 'cmdstanr')
plot(short_fit, nvariables = 10, variable = '^b_', regex = TRUE)
plot(short_fit, nvariables = 10, variable = '^sd', regex = TRUE)

# Full fit
full_fit <- brm(formula = formula_nl, data = dta, prior = prior_nl, iter = 1000, chains = 4, cores = 4, refresh = 0, backend = 'cmdstanr')
null_fit <- brm(formula = formulaNull_nl, data = dta, prior = priorNull_nl, iter = 1000, chains = 4, cores = 4, refresh = 0, backend = 'cmdstanr')
```

## Inspect the posterior fit

```{r inspect-fit, eval=FALSE}
full_sm <- summary(full_fit, robust = TRUE)
print(full_sm)

full_fix <- full_sm$fixed
full_fix_rn <- rownames(full_fix)

round(plogis(q = with(full_fix[full_fix_rn %in% 'LogitBase_Intercept', ], c(median = Estimate, percentile_2.5 = `l-95% CI`, percentile_97.5 = `u-95% CI`))), digits = 2)
round(plogis(q = with(full_fix[full_fix_rn %in% 'LogitLapse_Intercept', ], c(median = Estimate, percentile_2.5 = `l-95% CI`, percentile_97.5 = `u-95% CI`))), digits = 2)
round(exp(with(full_fix[full_fix_rn %in% 'LogWidth_Intercept', ], c(median = Estimate, percentile_2.5 = `l-95% CI`, percentile_97.5 = `u-95% CI`))), digits = 2)
round(with(full_fix[full_fix_rn %in% 'Inflex_Intercept', ], c(median = Estimate, percentile_2.5 = `l-95% CI`, percentile_97.5 = `u-95% CI`))), digits = 2)
round(with(full_fix[full_fix_rn %in% 'Inflex_typebeta', ], c(median = Estimate, percentile_2.5 = `l-95% CI`, percentile_97.5 = `u-95% CI`))), digits = 2)
```

## Model comparison and predictions

```{r nlgmm-comparison, eval=FALSE}
loo_full <- loo(full_fit)
loo_null <- loo(null_fit)
loo_compare(loo_full, loo_null)

full_cond <- brms::conditional_effects(full_fit, method = 'posterior_epred', cores = parallel::detectCores() - 1, effects = c('stimulus:type'))
pred_data <- full_cond$`stimulus:type`

plot(NULL, xlab = 'Stimulus intensity', ylab = 'Proportion correct', xlim = range(stimulus) * c(0.7, 1.3), ylim = c(0, 1))
abline(h = c(0, 1, base, 1 - lapse), lty = c(1, 1, 3, 3))
for (aa in unique(agg$animal)) {
  with(subset(agg, animal == aa), {
    points(x = stimulus, y = correct_incorrect, col = c('blue', 'orange3')[1 + type %in% levels(type)[2]], pch = c(20, 21)[1 + type %in% levels(type)[2]])
  })
}
with(subset(pred_data, type == 'beta'), {
  polygon(x = c(sort(stimulus), rev(sort(stimulus))), y = c(lower__[order(stimulus)], rev(upper__[order(stimulus)])), col = adjustcolor('orange', alpha.f = 50 / 256), border = NA)
})
with(subset(pred_data, type == 'alpha'), {
  polygon(x = c(sort(stimulus), rev(sort(stimulus))), y = c(lower__[order(stimulus)], rev(upper__[order(stimulus)])), col = adjustcolor('darkblue', alpha.f = 50 / 256), border = NA)
})
with(subset(pred_data, type == 'alpha'), lines(x = sort(stimulus), y = estimate__[order(stimulus)], col = 'darkblue', lwd = 5))
with(subset(pred_data, type == 'beta'), lines(x = sort(stimulus), y = estimate__[order(stimulus)], col = 'orange', lwd = 5))
legend('bottomright', legend = unique(agg$type), col = c('blue', 'orange3'), pch = c(20, 21))
```
