---
title: "Basic GLMM example"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
  pdf_document:
    toc: true
fig_caption: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
set.seed(17070523)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic GLMM example

This document simulates binomial data and fits a logistic mixed-effects model using `glmer()`.

## Simulate data

We generate response data for individual animals, stimuli, and two stimulus types.

```{r simulate-glmm}
n_animals <- 10
n_levels <- 7
n_types <- 2
replicates <- 20

pop_mean <- 5.00
pop_sd <- 1.00
slope_mean <- 3.00
slope_sd <- 1.03

type_mean <- 2.50
type_slope <- 3.5
type_animal_sd <- 0.55
extra_error_sd <- 0.10

animal <- rep(x = 1:n_animals, times = n_levels)
animal <- sort(animal)
animal <- rep(x = animal, times = n_types)
stimulus <- rep(x = 1:n_levels - 1, times = n_animals)
stimulus <- rep(x = stimulus, times = n_types)
stype <- rep(x = 1:n_types, times = n_animals * n_levels)
stype <- sort(stype)

animal <- rep(x = animal, times = replicates)
stimulus <- rep(x = stimulus, times = replicates)
stype <- rep(x = stype, times = replicates)

base_bias_animal <- rnorm(n = n_animals, mean = pop_mean, sd = pop_sd)
slope_animal <- rnorm(n = n_animals, mean = slope_mean, sd = slope_sd)
type_animal <- rnorm(n = n_animals, mean = 1.0, sd = type_animal_sd)

response_y <- rep(x = NA, length = n_animals * n_levels * n_types * replicates)
for (ii in seq_along(response_y)) {
  response_y[ii] <- base_bias_animal[animal[ii]] +
    (stype[ii] - 1) * type_mean +
    (slope_animal[animal[ii]] +
       (stype[ii] - 1) * type_slope * type_animal[animal[ii]]) * stimulus[ii] +
    rnorm(n = 1, mean = 0, sd = extra_error_sd)
}

correct_incorrect <- rbinom(n = length(response_y), size = 1, prob = plogis(q = response_y - mean(response_y)))

sim_data <- data.frame(
  correct_incorrect = correct_incorrect,
  animal = LETTERS[animal],
  stimulus = stimulus,
  type = factor(stype, labels = c('alpha', 'beta'))
)

summary(sim_data)
```

## Save and load data

```{r save-read-glmm}
root_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), 'Simulated Data')
file_name <- 'simulated_binomial_data.csv'
write.csv(file = file.path(root_dir, file_name), x = sim_data, row.names = FALSE)

dta <- read.csv(file = file.path(root_dir, file_name), header = TRUE)
dta <- within(dta, {
  animal <- factor(animal)
  type <- factor(type)
})

summary(dta)
```

## Plot the binomial data

We aggregate the proportion correct for each animal and stimulus level.

```{r plot-glmm, fig.width=7, fig.height=5}
clz <- sapply(length(unique(dta$animal)), colorRampPalette(c('red', 'blue', 'cyan4', 'magenta4', 'orange')))
clz <- sample(clz)

agg <- aggregate(correct_incorrect ~ stimulus * type * animal, data = dta, FUN = mean)
head(agg)

plot(NULL,
     xlab = 'Stimulus intensity',
     ylab = 'Proportion correct',
     xlim = range(stimulus) * c(0.7, 1.3),
     ylim = c(0, 1))
abline(h = c(0, 1))
for (aa in unique(agg$animal)) {
  with(subset(agg, animal == aa), {
    points(
      x = stimulus,
      y = correct_incorrect,
      col = clz[which(LETTERS %in% aa)],
      pch = c(20, 21)[1 + type %in% levels(type)[2]]
    )
  })
}
legend('topright', legend = unique(agg$animal), col = clz, pch = 20)
legend('bottomright', legend = unique(agg$type), pch = c(20, 21))
```

## Fit GLMMs

We fit a maximal logistic mixed-effects model and a null random-effects model.

```{r fit-glmm}
ctrl_opt <- glmerControl(optimizer = 'bobyqa')
glmm.max <- glmer(
  formula = correct_incorrect ~ stimulus * type + (1 + stimulus * type | animal),
  data = dta,
  control = ctrl_opt,
  family = binomial(link = 'logit')
)
glmm.null <- glmer(
  formula = correct_incorrect ~ 1 + (1 | animal),
  data = dta,
  family = binomial(link = 'logit')
)
```

## Model comparison

```{r compare-glmm}
extractAIC(glmm.max)[2]
extractAIC(glmm.null)[2]
anova(glmm.max, glmm.null, test = 'Chisq')
```

## Model reduction

We compare models with and without the interaction, stimulus, and type terms.

```{r glmm-model-reduction}
no_int <- update(glmm.max, . ~ . - stimulus:type - (1 + stimulus * type | animal) + (1 + stimulus + type | animal))
anova(glmm.max, no_int)

no_stimulus <- update(no_int, . ~ . - stimulus - (1 + stimulus + type | animal) + (1 + type | animal))
anova(no_int, no_stimulus)

no_type <- update(no_stimulus, . ~ . - type - (1 + type | animal) + (1 | animal))
anova(no_stimulus, no_type)
```

## Post-hoc comparisons

We use `emmeans` to compare overall type effects and slope differences.

```{r posthoc-glmm, message=FALSE}
library(emmeans)
library(pbkrtest)

emm_intercepts <- emmeans(glmm.max, specs = list(pairwise ~ type), adjust = 'sidak')
summary(emm_intercepts)

emm_slopes_interact <- emtrends(glmm.max, specs = list(pairwise ~ type), var = 'stimulus', adjust = 'sidak')
summary(emm_slopes_interact)
```

## Prediction plots

We compute predicted values and plot the fitted response curves for each type.

```{r glmm-predictions, fig.width=7, fig.height=5}
if (Sys.info()[['sysname']] == 'Windows') {
  require(snow)
}
newdta <- with(dta, expand.grid(stimulus = seq(min(stimulus), max(stimulus), length.out = 20), type = unique(type), animal = unique(animal)))
prd <- predict(glmm.max, newdata = newdta, type = 'response')

pfun <- function(x) predict(x, newdata = newdta, type = 'response')
avail.cores <- parallel::detectCores() - 1
clt <- parallel::makeCluster(spec = avail.cores, type = if (Sys.info()[['sysname']] == 'Windows') 'SOCK' else 'FORK')
parallel::clusterExport(clt, list('glmm.max', 'dta', 'newdta'))

bt <- bootMer(
  glmm.max,
  FUN = pfun,
  nsim = 50,
  re.form = NULL,
  parallel = ifelse(Sys.info()[['sysname']] == 'Windows', 'snow', 'multicore'),
  ncpus = parallel::detectCores() - 1,
  cl = clt
)
parallel::stopCluster(clt)

pred_q <- aggregate(pred ~ stimulus * type, data = with(bt, data.frame(newdta, pred = c(t(t)))), FUN = quantile, probs = c(0.025, 0.975))
mod_mean <- aggregate(prd ~ stimulus * type, data = cbind(newdta, prd), FUN = function(x) plogis(mean(qlogis(x))))
param_data <- merge(merge(newdta, pred_q, all = TRUE), mod_mean)
mod_pred <- within(param_data, {
  mod_mean <- prd
  CI_02.5 <- pred[, '2.5%']
  CI_97.5 <- pred[, '97.5%']
  rm(list = c('pred', 'prd'))
})

agg <- aggregate(correct_incorrect ~ stimulus * type * animal, data = dta, FUN = mean)
plot(NULL,
     xlab = 'Stimulus intensity',
     ylab = 'Proportion correct',
     xlim = range(stimulus) * c(0.7, 1.3),
     ylim = c(0, 1))
abline(h = c(0, 1))
for (aa in unique(agg$animal)) {
  with(subset(agg, animal == aa), {
    points(x = stimulus, y = correct_incorrect, col = c('orange2', 'darkblue')[1 + type %in% 'alpha'], pch = 21)
  })
}
with(subset(mod_pred, type == 'alpha'), {
  polygon(x = c(sort(stimulus), sort(stimulus, decreasing = TRUE)), y = c(CI_02.5[order(stimulus)], CI_97.5[order(stimulus, decreasing = TRUE)]), col = adjustcolor('darkblue', alpha.f = 0.2), border = NA)
  lines(sort(stimulus), mod_mean[order(stimulus)], col = 'darkblue', lwd = 3)
})
with(subset(mod_pred, type == 'beta'), {
  polygon(x = c(sort(stimulus), sort(stimulus, decreasing = TRUE)), y = c(CI_02.5[order(stimulus)], CI_97.5[order(stimulus, decreasing = TRUE)]), col = adjustcolor('orange2', alpha.f = 0.2), border = NA)
  lines(sort(stimulus), mod_mean[order(stimulus)], col = 'orange2', lwd = 3)
})
```
