---
title: "Hypothesis testing example"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
  pdf_document:
    toc: true
    fig_caption: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE}
library(report)
set.seed(20220622)
knitr::opts_chunk$set(echo = TRUE)
```

# Hypothesis testing example

This example demonstrates two-group hypothesis tests using base R functions like `t.test` and `wilcox.test`. 
It also uses the `report` package for easy reporting of results.

## Enter data

We will simulate random data for two experimental conditions, A and B.

```{r simulate-data}
dt_A <- rnorm(n = 10, mean = 1.0, sd = 0.3)
fct_A <- rep(x = 'A', times = length(dt_A))

dt_B <- rnorm(n = 10, mean = 1.5, sd = 0.3)
fct_B <- rep(x = 'B', times = length(dt_B))

dt <- data.frame(
  value = c(dt_A, dt_B),
  condition = c(fct_A, fct_B)
)

summary(dt)
```

## Inspect the data

We will visualise the distribution with a boxplot and raw data points.

```{r inspect-data, fig.width=6, fig.height=4}
boxplot(value ~ condition,
        data = dt,
        ylab = 'body mass (g)',
        xlab = 'condition',
        ylim = range(dt$value),
        pars = list(boxwex = 0.5),
        outline = FALSE,
        col = 'gray95',
        cex.lab = 1.5)

stripchart(value ~ condition,
           data = dt,
           vertical = TRUE,
           add = TRUE,
           method = 'jitter',
           pch = 21,
           bg = 'darkblue',
           col = 'white',
           cex = 1.5)
```

## Perform a Student's t-test

We ask whether `condition` predicts `value`, using a two-sided Student's t-test with equal variances.

```{r t-test}
tt <- t.test(
  formula = value ~ condition,
  data = dt,
  alternative = "two.sided",
  var.equal = TRUE
)

print(tt)

suppressWarnings(report(tt))
```

## Welch's t-test for unequal variances

Now we simulate a dataset where the two groups have different variances and apply Welch's t-test.

```{r simulate-unequal}
dt_unequal <- within(dt, {
  value[condition %in% 'B'] <- rnorm(n = 10, mean = 1.5, sd = 0.15)
})

summary(dt_unequal)

tt_welch <- t.test(
  formula = value ~ condition,
  data = dt_unequal,
  alternative = "two.sided",
  var.equal = FALSE
)

print(tt_welch)

suppressWarnings(report(tt_welch))
```

# Non-normal distributions and the Mann-Whitney test

For data that may violate the normality assumption, we simulate an asymmetric distribution and compare groups using Welch's t-test and the Mann-Whitney rank sum test.

```{r simulate-nonnormal}
dt_un_A <- rnorm(n = 10, mean = 1.0, sd = 0.3) + rpois(n = 10, lambda = 3.4)
fct_un_A <- rep(x = 'A', times = length(dt_un_A))

dt_un_B <- rnorm(n = 10, mean = 3.5, sd = 0.6) + rpois(n = 10, lambda = 3.4)
fct_un_B <- rep(x = 'B', times = length(dt_un_B))

dt_unknown <- data.frame(
  value = c(dt_un_A, dt_un_B),
  condition = c(fct_un_A, fct_un_B)
)

summary(dt_unknown)
```

```{r inspect-nonnormal, fig.width=6, fig.height=4}
boxplot(value ~ condition,
        data = dt_unknown,
        ylab = 'body mass (g)',
        xlab = 'condition',
        ylim = range(dt_unknown$value),
        pars = list(boxwex = 0.5),
        outline = FALSE,
        col = 'gray95',
        cex.lab = 1.5)

stripchart(value ~ condition,
           data = dt_unknown,
           vertical = TRUE,
           add = TRUE,
           method = 'jitter',
           pch = 21,
           bg = 'darkblue',
           col = 'white',
           cex = 1.5)
```

```{r mean-ci}
mean_unknown <- aggregate(
  value ~ condition,
  data = dt_unknown,
  FUN = function(x) {
    sum_stat <- mean(x) + 1.96 * c(0, -1, 1) * sd(x) / sqrt(length(x))
    names(sum_stat) <- c('mean', 'lower CI', 'upper CI')
    return(sum_stat)
  }
)

print(mean_unknown)
```

```{r t-test-unknown}
tt_unknown <- t.test(
  formula = value ~ condition,
  data = dt_unknown,
  alternative = "two.sided",
  var.equal = FALSE
)

print(tt_unknown)

mww <- wilcox.test(
  value ~ condition,
  data = dt_unknown,
  alternative = "two.sided"
)

print(mww)
```

## Rank-transformed visualisation

The plot below shows both the original values and the rank-transformed values for the two conditions.

```{r rank-plot, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
with(dt_unknown, {
  hist(value[condition %in% 'A'],
       breaks = 1e2,
       col = 'blue',
       xlim = range(value),
       xlab = 'value',
       main = 'original values')
  hist(value[condition %in% 'B'],
       breaks = 1e2,
       col = 'red',
       add = TRUE)
  hist(rank(value)[condition %in% 'A'],
       breaks = 1e2,
       col = 'blue',
       xlim = c(0, max(rank(value))),
       xlab = 'value',
       main = 'rank-transformed')
  hist(rank(value)[condition %in% 'B'],
       breaks = 1e2,
       col = 'red',
       add = TRUE)
})
```
