---
title: "ANOVA example"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
  pdf_document:
    toc: true
    fig_caption: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE}
library(report)
set.seed(20220622)
knitr::opts_chunk$set(echo = TRUE)
```

# ANOVA example

This document demonstrates a simple one-way ANOVA example using simulated data.

## Enter data

We simulate two groups and then add a third group for the ANOVA.

```{r simulate-data}
dt_A <- rnorm(n = 10, mean = 1.0, sd = 0.3)
fct_A <- rep(x = 'A', times = length(dt_A))

dt_B <- rnorm(n = 10, mean = 1.5, sd = 0.3)
fct_B <- rep(x = 'B', times = length(dt_B))

dt <- data.frame(
  value = c(dt_A, dt_B),
  condition = c(fct_A, fct_B)
)

summary(dt)
```

## Perform a t-test

We begin with a two-sample Student's t-test on the two initial conditions.

```{r t-test}
tt <- t.test(
  formula = value ~ condition,
  data = dt,
  alternative = "two.sided",
  var.equal = TRUE
)

print(tt)
suppressWarnings(report(tt))
```

## Add another condition for ANOVA

We add a third condition and merge it into the dataset so that ANOVA can be used.

```{r add-condition}
dt_C <- rnorm(n = 10, mean = 1.1, sd = 0.3)
fct_C <- rep(x = 'C', times = length(dt_C))

dt_new <- data.frame(value = dt_C, condition = fct_C)

dt_all <- merge.data.frame(x = dt, y = dt_new, all = TRUE)

summary(dt_all)
```

## One-way ANOVA using aov()

Because there are now three conditions, we use `aov()` instead of `t.test()`.

```{r anova}
av <- aov(value ~ condition, data = dt_all)
av_summary <- summary(av)
print(av_summary)

av_Ftest <- anova(object = av, test = 'F')
print(av_Ftest)

suppressWarnings(report(av))
```

## ANOVA post-hoc tests with correction

We use `emmeans` to perform pairwise comparisons with Sidak correction.

```{r post-hoc, message=FALSE}
library(emmeans)
av_em <- emmeans(object = av, specs = pairwise ~ condition, adjust = 'sidak')
av_ph <- emmeans::test(object = av_em)
print(av_ph$contrasts)
```

## Use a linear model instead of ANOVA

This section shows how the same question can be answered through `lm()`.

```{r linear-model}
lmd <- lm(value ~ condition, data = dt_all)
lmd_Ftest <- anova(object = lmd, test = 'F')
print(lmd_Ftest)

suppressWarnings(report(lmd_Ftest))

lmd_summary <- summary(lmd)
print(lmd_summary)
```

## Post-hoc tests on the linear model

We perform the same pairwise comparisons using the linear model object.

```{r linear-model-posthoc, message=FALSE}
lmd_em <- emmeans(object = lmd, specs = pairwise ~ condition, adjust = 'sidak')
lmd_ph <- emmeans::test(object = lmd_em)
print(lmd_ph$contrast)
```
