---
title: "Basic LMM example"
author: "James Foster"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
  pdf_document:
    toc: true
    fig_caption: true
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
set.seed(17070523)
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Linear Mixed Model example

This example simulates a repeated-measures experiment and fits mixed-effects models using `lmer()`.

## Simulate some data

We simulate responses for multiple animals, stimulus levels, and stimulus types.

```{r simulate-lmm}
n_animals <- 10
n_levels <- 7
n_types <- 2
replicates <- 20

pop_mean <- 5.00
pop_sd <- 1.00
slope_mean <- 3.00
slope_sd <- 1.03

type_mean <- 2.50
type_slope <- 3.5
type_animal_sd <- 0.55
extra_error_sd <- 0.10

animal <- rep(x = 1:n_animals, times = n_levels)
animal <- sort(animal)
animal <- rep(x = animal, times = n_types)
stimulus <- rep(x = 1:n_levels - 1, times = n_animals)
stimulus <- rep(x = stimulus, times = n_types)
stype <- rep(x = 1:n_types, times = n_animals * n_levels)
stype <- sort(stype)

animal <- rep(x = animal, times = replicates)
stimulus <- rep(x = stimulus, times = replicates)
stype <- rep(x = stype, times = replicates)

base_bias_animal <- rnorm(n = n_animals, mean = pop_mean, sd = pop_sd)
slope_animal <- rnorm(n = n_animals, mean = slope_mean, sd = slope_sd)
type_animal <- rnorm(n = n_animals, mean = 1.0, sd = type_animal_sd)

response_y <- rep(x = NA, length = n_animals * n_levels * n_types * replicates)
for (ii in seq_along(response_y)) {
  response_y[ii] <- base_bias_animal[animal[ii]] +
    (stype[ii] - 1) * type_mean +
    (slope_animal[animal[ii]] +
       (stype[ii] - 1) * type_slope * type_animal[animal[ii]]) * stimulus[ii] +
    rnorm(n = 1, mean = 0, sd = extra_error_sd)
}

sim_data <- data.frame(
  response_y = response_y,
  animal = LETTERS[animal],
  stimulus = stimulus,
  type = factor(stype, labels = c('alpha', 'beta'))
)

summary(sim_data)
```

## Save and reload data

We write the data to CSV and reload it for analysis, then format categorical variables.

```{r save-read-data}
root_dir <- file.path(dirname(knitr::current_input(dir = TRUE)), 'Simulated Data')
file_name <- 'simulated_data.csv'
write.csv(file = file.path(root_dir, file_name), x = sim_data, row.names = FALSE)

dta <- read.csv(file = file.path(root_dir, file_name), header = TRUE)
dta <- within(dta, {
  animal <- factor(animal)
  type <- factor(type)
})

summary(dta)
```

## Plot the data

We visualise the individual animal responses by stimulus and type.

```{r plot-lmm, fig.width=7, fig.height=5}
clz <- sapply(length(unique(dta$animal)), colorRampPalette(c('red', 'blue', 'cyan4', 'magenta4', 'orange')))
clz <- sample(clz)

plot(NULL,
     xlab = 'Stimulus intensity',
     ylab = 'Response strength',
     xlim = range(stimulus) * c(0.7, 1.3),
     ylim = range(response_y) * c(0.7, 1.3))
for (aa in unique(dta$animal)) {
  with(subset(dta, animal == aa), {
    points(
      x = stimulus,
      y = response_y,
      col = clz[which(LETTERS %in% aa)],
      pch = c(20, 21)[1 + type %in% levels(type)[2]]
    )
  })
}
legend('topright', legend = unique(dta$animal), col = clz, pch = 20, cex = 0.7)
legend('bottomright', legend = unique(dta$type), pch = c(20, 21), cex = 0.7)
```

## Fit the maximal model

We fit a maximal linear mixed model with random intercepts and slopes for each animal.

```{r fit-maximal,library=FALSE}
ctrl_opt <- lmerControl(optimizer = 'bobyqa')
mixmod.max <- lmer(
  formula = response_y ~ stimulus * type + (1 + stimulus * type | animal),
  data = dta,
  control = ctrl_opt
)
mixmod.null <- lmer(formula = response_y ~ 1 + (1 | animal), data = dta)
```

## Compare models

We compare the maximal model to the null model using AIC and a likelihood ratio test.

```{r model-comparison}
extractAIC(mixmod.max)[2]
extractAIC(mixmod.null)[2]
anova(mixmod.max, mixmod.null, test = 'Chisq')
```

## Inspect the maximal model ANOVA

```{r inspect-maximal}
anova(mixmod.max)
```

## Model reduction

We test whether interaction, stimulus, or type can be removed without worsening model fit.

```{r model-reduction}
no_int <- update(mixmod.max, . ~ . - stimulus:type - 
                   (1 + stimulus * type | animal) + 
                   (1 + stimulus + type | animal))
anova(mixmod.max, no_int)

no_stimulus <- update(no_int, . ~ . - stimulus - 
                        (1 + stimulus + type | animal) + 
                        (1 + type | animal))
anova(no_int, no_stimulus)

no_type <- update(no_stimulus, . ~ . - type - 
                    (1 + type | animal) + (1 | animal))
anova(no_stimulus, no_type)
```

## Post-hoc comparisons

We use `emmeans` and `pbkrtest` to make pairwise comparisons for type and slope effects.

```{r posthoc-lmm, message=FALSE}
library(emmeans)
library(pbkrtest)

emm_intercepts <- emmeans(mixmod.max, specs = list(pairwise ~ type), 
                          adjust = 'sidak')
summary(emm_intercepts)

emm_slopes_interact <- emtrends(mixmod.max, specs = list(pairwise ~ type), 
                                var = 'stimulus', adjust = 'sidak')
summary(emm_slopes_interact)
```

## Prediction plots

We generate fitted predictions and plot the model with confidence intervals.

```{r plot-predictions, fig.width=7, fig.height=5}
newdta <- with(dta, expand.grid(stimulus = unique(stimulus), 
                                type = unique(type), 
                                animal = unique(animal)))
prd <- predict(mixmod.max, newdata = newdta)

pfun <- function(x) predict(x, newdata = newdta)
avail.cores <- parallel::detectCores() - 1
clt <- parallel::makeCluster(spec = avail.cores, type = 'PSOCK')
parallel::clusterExport(clt, list('mixmod.max', 'dta', 'newdta'))

bt <- bootMer(
  mixmod.max,
  FUN = pfun,
  nsim = 100,
  re.form = NA,
  parallel = 'snow',
  ncpus = parallel::detectCores() - 1,
  cl = clt
)
parallel::stopCluster(clt)

param_q <- aggregate(pred ~ stimulus * type, 
                     data = with(bt, data.frame(newdta, pred = c(t(t)))), 
                     FUN = quantile, 
                     probs = c(0.025, 0.975))
mod_mean <- aggregate(prd ~ stimulus * type, 
                      data = cbind(newdta, prd), 
                      FUN = mean)
param_data <- merge(merge(newdta, param_q, all = TRUE), mod_mean)
mod_pred <- within(param_data, {
  mod_mean <- prd
  CI_02.5 <- pred[, '2.5%']
  CI_97.5 <- pred[, '97.5%']
  rm(list = c('pred', 'prd'))
})

with(dta, plot(x = stimulus, y = response_y, 
               bg = adjustcolor(col = c('orange2', 'darkblue')[
                 1 + type %in% 'alpha'], alpha.f = 0.5), 
               col = 'black', pch = 21))
with(subset(mod_pred, type %in% 'alpha'), {
  polygon(x = c(sort(stimulus), sort(stimulus, decreasing = TRUE)),
          y = c(CI_02.5[order(stimulus)], 
                CI_97.5[order(stimulus, decreasing = TRUE)]),
          col = adjustcolor('darkblue', alpha.f = 0.2), border = NA)
  lines(sort(stimulus), 
        mod_mean[order(stimulus)], 
        col = 'darkblue', lwd = 3)
})
with(subset(mod_pred, type %in% 'beta'), {
  polygon(x = c(sort(stimulus), sort(stimulus, decreasing = TRUE)),
          y = c(CI_02.5[order(stimulus)], 
                CI_97.5[order(stimulus, decreasing = TRUE)]),
          col = adjustcolor('orange2', alpha.f = 0.2), border = NA)
  lines(sort(stimulus),
        mod_mean[order(stimulus)], 
        col = 'orange2', lwd = 3)
})
```
